---
title: "Does cost recovery lead to asset growth for electric utilities?"
author: "Edgar Aguilar"
institute: "American University"
date: "ea9339a@american.edu"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [robot, robot-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
---  

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  comment   = NA,
  dev       = "svglite",
  fig.align = "center",
  fig.width = 7,
  fig.height= 5,
  out.width = "80%"
)

# Load required packages
library(readxl)
library(ggplot2)
library(dplyr)
library(lfe)
library(modelsummary)
library(scales)

# Load and prepare the dataset
data_file <- "slides_utilities_data.xlsx"
if (!file.exists(data_file)) {
  stop("Data file 'slides_utilities_data.xlsx' not found.")
}
data_slides <- read_xlsx(data_file) %>%
  rename(IncomeGroup = `Income Group`) %>%
  mutate(
    `Average Cost Recovery`    = as.numeric(`Average Cost Recovery`),
    `Total Assets Growth Rate` = as.numeric(`Total Assets Growth Rate`),
    ACR_percent   = `Average Cost Recovery` * 100,
    TAGR_percent  = `Total Assets Growth Rate` * 100
  )
```

# Overview
- Electric utilities are essential to expanding and maintaining reliable electricity access.

- Asset growth reflects infrastructure expansion and improved service capacity.

- This analysis tests whether utilities that recover their costs are better positioned to grow their assets.

- I begin with descriptive statistics, then test the cost recovery–growth link using regression.
	
---
# Description of the dataset
- 185 utilities across 93 countries.

- Majority are from lower- and upper-middle income countries.

- Two indicators are central to this analysis:
    - Average Cost Recovery (ACR): 2012–2022
    - Total Asset Growth Rate (TAGR): Annualized over 2012–2022
	
---
# Distribution of cost recovery
- Wide variation in cost recovery.

- 116 utilities fall below the 100% benchmark (shown in red dashed line).

- 69 utilities exceed 100% recovery.
```{r cost-recovery-histogram, fig.width=7, fig.height=3, echo=FALSE}
ggplot(data_slides, aes(x = ACR_percent)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", boundary = 0) +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red", size = 1) +
  scale_x_continuous(
    breaks = seq(0, 200, 25),
    limits = c(0, 200),
    labels = percent_format(scale = 1)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)) # no gap at bottom, tiny gap at top
  ) +
  labs(
    title = "Histogram of Average Cost Recovery",
    x     = "Cost Recovery (%)",
    y     = "Count of Utilities"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),   # bottom line only
    axis.line.y = element_blank(),                 # no vertical axis line
    plot.title  = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.x  = element_text(face = "bold"),
    axis.text.y  = element_text(face = "bold")
  )
```

---
# Asset growth vs. cost recovery
- The scatterplot shows a weak and slightly negative relationship between cost recovery and asset growth.

- Regression analysis confirms that cost recovery is not a statistically significant predictor of asset growth (p = 0.22).

- This suggests that higher cost recovery alone is not sufficient to guarantee infrastructure expansion.
.pull-left[
```{r scatter-plot, fig.width=10, fig.height=7, echo=FALSE}
ggplot(data_slides, aes(x = ACR_percent, y = TAGR_percent)) +
  geom_point(alpha = 0.7, size = 3.5) +  # larger points
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black", size = 1.2) +
  scale_x_continuous(
    limits = c(0, 200),
    breaks = seq(0, 200, 25),
    labels = scales::percent_format(scale = 1)
  ) +
  scale_y_continuous(
    limits = c(0, 40),
    breaks = seq(0, 40, 10),
    labels = scales::percent_format(scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Asset Growth vs. Cost Recovery",
    x     = "Cost Recovery (%)",
    y     = "Asset Growth Rate (%)"
  ) +
  theme_minimal(base_size = 20) +  # even bigger base size
  theme(
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.line.x        = element_line(color = "black"),
    axis.line.y        = element_blank(),
    plot.title         = element_text(hjust = 0.5, face = "bold", size = 26),
    axis.title.x       = element_text(face = "bold", size = 22),
    axis.title.y       = element_text(face = "bold", size = 22),
    axis.text.x        = element_text(face = "bold", size = 20),
    axis.text.y        = element_text(face = "bold", size = 20),
    legend.title       = element_text(face = "bold", size = 20),
    legend.text        = element_text(size = 18),
    legend.position    = "top"
  )
```
]
.pull-right[
```{r regression-table, echo=FALSE}
# Run the regression
model <- lm(`Total Assets Growth Rate` ~ `Average Cost Recovery`, data = data_slides)

# Extract regression coefficients and rename columns
reg_table <- as.data.frame(summary(model)$coefficients)
colnames(reg_table) <- c("Estimate", "Std. Error", "t-value", "p-value")

# Print the table using kableExtra with a chosen font size
library(knitr)
library(kableExtra)
kable(reg_table, caption = "Regression Results: Total Assets Growth Rate on Average Cost Recovery", align = "c") %>%
  kable_styling(full_width = FALSE, font_size = 10)
```
]

---
# Cost Recovery by Income Group
.pull-left[
- Clear gradient observed: cost recovery is generally lower in low-income countries and higher in upper-middle and high-income countries.

- Variability in cost recovery is largest among lower-middle income utilities.

- These patterns suggest that broader economic and institutional factors, linked to country income levels, influence utilities’ financial performance and investment capacity.

]

.pull-right[
```{r boxplot-income, fig.width=7, fig.height=5, echo=FALSE}
# First, reorder IncomeGroup as a factor
data_slides <- data_slides %>%
  mutate(IncomeGroup = factor(IncomeGroup, levels = c(
    "Low income", "Lower middle income", "Upper middle income", "High income"
  )))

# Now plot
ggplot(data_slides, aes(x = IncomeGroup, y = ACR_percent)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.5, outlier.size = 1.5) +
  
  # Add mean and median with manual color legend
  stat_summary(
    aes(color = "Mean"), fun = mean, geom = "point", shape = 19, size = 2
  ) +
  stat_summary(
    aes(color = "Median"), fun = median, geom = "point", shape = 17, size = 2
  ) +
  
  scale_y_continuous(
    limits = c(0, 200),
    labels = percent_format(scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  
  # Define color legend manually
  scale_color_manual(
    name = "Statistics",
    values = c("Mean" = "red", "Median" = "blue")
  ) +
  
  labs(
    title = "Cost Recovery by Income Group",
    x     = "Income Group",
    y     = "Cost Recovery (%)"
  ) +
  
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_blank(),
    axis.text.x = element_text(face = "bold", size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(face = "bold", size = 10),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 12),
    legend.text  = element_text(size = 10),
    legend.position = "top"
  )
```
]
---

# Thank you!

